---
title: "Cyclistic_2023_analysis"
author: "Atharv Verma"
date: "`r Sys.Date()`"
output: html_document
---

# Cyclistic Bike-Share 2023 Analysis

### Step-1: Data Gathering

The preprocessed data is stored in overall_data_2023.csv file from the database.

#### Loading libraries

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

#### Importing data:

```{r}
tripdata <- read.csv("overall_data_2023.csv", header = TRUE, sep = ",")
tripdata
```

### Step-2: Statistics

#### Summary Stats:

summary_stats shows the general stats of the data.

```{r}
mode_start_day <- tripdata %>%
  count(start_day) %>%
  filter(n == max(n)) %>%
  pull(start_day)

summary_stats <- tripdata %>% 
  summarise(mean_duration = mean(duration, na.rm = TRUE),
            max_duration = max(duration, na.rm = TRUE),
            total_casual_riders = sum(membership_type == "casual", na.rm = TRUE),
            total_member_riders = sum(membership_type == "member", na.rm = TRUE),
            mode_start_day = mode_start_day)

print(summary_stats)
```

-   Average ride duration: 18 minutes approx.
-   Max ride duration: 98489 minutes
-   Total Casual Rides in 2023: 4,078,350
-   Total Member Rides in 2023: 7,171,103
-   Most frequent rides on: Saturday

#### Duration Stats:

Statistics of ride duration for casual and member riders.

```{r}
duration_stats <- tripdata %>% 
  group_by(membership_type) %>% 
  summarise(mean_duration = mean(duration, na.rm = TRUE),
            max_duration = max(duration, na.rm = TRUE),
            median_duration = median(duration, na.rm = TRUE),
            total_riders = n())
print(duration_stats)
```

#### Day Stats:

Statistics of start day for casual and member riders.

```{r}
day_stats <- tripdata %>% 
  mutate(start_day = str_trim(start_day)) %>% 
  group_by(membership_type, start_day) %>% 
  summarise(total_rides = n()) %>% 
  mutate(start_day = factor(start_day, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>% 
  arrange(membership_type, start_day)
print(day_stats)
```

#### Time of Day Stats:

Statistics of time of day for casual and member riders.

```{r}
time_of_day_stats <- tripdata %>% 
  group_by(membership_type, time_of_day) %>% 
  summarise(total_rides = n()) %>% 
  arrange(membership_type, time_of_day)
print(time_of_day_stats)
```

#### Bike Type Stats:

Statistics of bike type used by casual and member riders.

```{r}
bike_type_stats <- tripdata %>% 
  group_by(membership_type, bike_type) %>% 
  summarise(total_rides = n()) %>% 
  arrange(membership_type, -total_rides)
print(bike_type_stats)
```

#### Month Stats:

Statistics of Start month of rides by casual and member riders.

```{r}
month_stats <- tripdata %>% 
  mutate(start_month = str_trim(start_month)) %>% 
  group_by(membership_type, start_month) %>% 
  summarise(total_rides = n()) %>% 
  mutate(start_month = factor(start_month, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))) %>%
  arrange(start_month)
print(month_stats)
```

#### Hour Stats:

Statistics of Start hour of rides by casual and member riders.

```{r}
hour_stats <- tripdata %>%
  group_by(membership_type, start_hour) %>%
  summarise(total_rides = n(), .groups = 'drop') %>% 
  mutate(start_hour = factor(start_hour, 
                             levels = c("12 AM", "1 AM", "2 AM", "3 AM", "4 AM", "5 AM", "6 AM", "7 AM", 
                                        "8 AM", "9 AM", "10 AM", "11 AM",
                                        "12 PM", "1 PM", "2 PM", "3 PM", "4 PM", "5 PM", "6 PM", "7 PM", 
                                        "8 PM", "9 PM", "10 PM", "11 PM"))) %>%
  arrange(membership_type, start_hour)
print(hour_stats)
```

#### Season Stats:

Statistics of season of rides by casual and member riders.

```{r}
season_membership_stats <- tripdata %>% 
  group_by(membership_type, season) %>% 
  summarise(total_rides = n())
print(season_membership_stats)
```

#### Season Percentage Stats:

Statistics of season percentage rides of total riders.

```{r}
season_stats <- tripdata %>% 
  group_by(season) %>% 
  summarise(total_rides = n()) %>% 
  mutate(percentage = total_rides / sum(total_rides) * 100)  # Calculate percentage
print(season_stats)
```

### Step-3: Visualization and Insights

#### Ride Duration by Membership Type

```{r pressure, echo=FALSE}
ggplot(tripdata, aes(x = membership_type, y=duration)) + geom_boxplot() + labs(title = "Trip Duration by Membership Type", y= "Duration(minutes)", x="Membership Type")
```

**Insights:**

-   Casual riders ride for longer duration.

-   Casual ride duration is scattered and varying.

-   Member riders ride for shorter duration.

-   Member ride duration is fairly consistent for majority of riders.

#### Ride by Days of Week

```{r}
ggplot(day_stats, aes(x = start_day, y=total_rides, fill = membership_type)) + geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Rides by Day of the Week", x = "Day of the Week", y = "Total Rides")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Insights:**

-   Casual rides are almost half of Member rides during weekdays.

-   Casual rides almost catch upto Member rides on weekends.

-   Casual riders use more rides on weekends.

-   Member riders use more rides on weekdays.

#### Ride by Time of Day

```{r}
#Rides by Time of Day
ggplot(time_of_day_stats, aes(x = time_of_day, y=total_rides, fill = membership_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Rides by Time of Day", x = "Time of Day", y="Total Rides")
```

**Insights:**

-   Both casual and member riders use bike rides during Afternoon hours the most.

-   Total rides are the lowest during Night hours for both casual and member riders.

#### Bike Type Usage by Membership Type

```{r}
#Bike Type Usage
library(scales)
ggplot(bike_type_stats, aes(x = bike_type, y = total_rides, fill = membership_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Bike Type Usage by Membership Type", 
    x = "Bike Type", 
    y="Total Rides"
    ) + 
  scale_y_continuous(labels = scales::comma)
```

**Insights:**

-   Docked bikes are only booked by Casual riders.

-   Member riders do not book docked bikes.

#### Rides by Month

```{r}
ggplot(month_stats, aes(x = start_month, y = total_rides, fill = membership_type)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Total Rides by Month", x = "Month", y = "Total Rides") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Insights:**

-   Hot months like June, July, and August are the busiest.

#### Rides by Hour

```{r}
ggplot(hour_stats, aes(x = start_hour, y = total_rides, color = membership_type, group = membership_type)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total Rides by Hour",
    x = "Start Hour",
    y = "Total Rides",
    color = "Membership Type"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Insights:**

-   Casual rides peaks till 5PM.

-   Member rides correlates to the standard office hours.

#### Rides by Season

```{r}
# Create a bar chart
ggplot(season_membership_stats, aes(x = season, y = total_rides, fill = membership_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Position bars side by side
  labs(
    title = "Total Rides by Season and Membership Type",
    x = "Season",
    y = "Total Rides",
    fill = "Membership Type"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +  # Use a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5)  # Center title
  )

season_stats <- tripdata %>% 
  group_by(season) %>% 
  summarise(total_rides = n()) %>% 
  mutate(percentage = total_rides / sum(total_rides) * 100)  # Calculate percentage
print(season_stats)

# Create a donut chart
ggplot(season_stats, aes(x = "", y = total_rides, fill = season)) +
  geom_bar(stat = "identity", width = 0.5, position = "stack", color = "white") +  # Create bars with white border
  coord_polar("y") +  # Convert to polar coordinates
  labs(
    title = "Percentage of Total Rides by Season",
    fill = "Season"
  ) +
  theme_void() +  # Remove axes and grid
  theme(
    plot.title = element_text(hjust = 0.5),  # Center title
    legend.position = "right"  # Position legend
  ) +
  # Create the hole for the donut
  geom_polygon(data = data.frame(x = 0, y = 0, r = 0.4), 
               aes(x = x, y = y), fill = "white", inherit.aes = FALSE) +  # Create the donut hole
  # Add percentage labels on the donut segments
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 5, fontface = "bold")  # Adjust text size and style
```

**Insights:**

-   Summer months are the busiest.

-   Massive fall-out in Winter months.
